---
title: "Lecture_2"
author: "Drew Steen"
date: "1/26/2022"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error = TRUE)
library(tidyverse)
```

# Most important data classes

## (Atomic) vectors

Talked about these in Lecture 1.5. We can create a vector with:

* `vector()`
* `c()`
* `seq()`
* `:`

```{r}
a <- c(1, 2, 3)
print(a)
```

```{r}
b <- seq(from = 1, to = 3, by = 1)
print(b)
```

```{r}
c <- 1:3 # "infix" operator, because it goes between numbers
```

### Vector coercion

```{r}
a <- c(1, "two", 3)
print(a)
```

#### Data types

* logical: TRUE/FALSE
* integer: 1, 2, 3
* double: 1.00000..., 3.14159 (32 bits of precision, scientific notation)
* character
* factor

R *coerces* data types to the *least general* type that works

```{r}
my.logical <- c(TRUE, FALSE, TRUE)
class(my.logical)
```

```{r}
my.result <- my.logical + 3
class(my.result)
```

#### Factors

Factors *look* like character vectors but are actually named itegers.
```{r}
my.fac <- factor(1:3)
```

### Accessing vector elements

We can access by *index*. Note R indices start with 1.

```{r}
my.vec <- c("a", "b", "c") # note the built-in vector letters is useful
my.vec[c(1,3)] # note that this expression is only 2 elements long
```

We can also access with a logical vector. This is particularly useful in selecting based on a test.

```{r}
my.vec[c(TRUE, FALSE, TRUE)] # this logical vector has the same length as the vector
```

```{r}
str_detect(my.vec, "[ac]") # this tests for a OR c - str_detect is short for "string detect"
```

```{r}
my.vec[str_detect(my.vec, "[ac]")]
```

```{r}
num.vec <- 1:15
num.vec[num.vec >= 8]
```

Remember that we can name the elements of a vector, and use those names to return values
```{r}
named.vec <- c(first = 1, second = 2, third = 3)
named.vec["second"]
```

# Lists

Lists store different kinds of information that need to be held together in one place. I think of them like Santa's bag - you can put anything in a list.

```{r}
my_list <- list(dataset = mtcars, my.favorite.letter = "f", some.numbers = 1:7)
summary(my_list)
```

Access list elements by element number:

```{r}
my_list[2]
```

```{r}
my_list[3] + 3
```
This didn't work because `my_list[3]` returns a list containing the vector we want as its only element, rather than the vector itself. We can get the actual list with double brackets, `[[`:

```{r}
my_list[[3]] + 3
```

`[[` gives us the actual list element, `[` gives us a list containing only the element that we asked for.

Access with `$`
```{r}
my_list$some.numbers + 3
```

Access by name
```{r}
my_list[["some.numbers"]] + 3
```

An example where accessing by name is really useful:

```{r}
model <- lm(mpg ~ disp, data = mtcars) # linear model
model$coefficients
```

## Data frames

Data frames are a special kind of list, with two extra rules:

* Each element is an atomic vector^[There is an exception: it is possible to create a "list-column" in a data frame. This is really useful, but we won't have time to talk about it here. More great information is [here](https://jennybc.github.io/purrr-tutorial/ls13_list-columns.html).]
* Each vector must be the same length.

Thus, a data frame is a bit like an Excel spreadsheet. Typically we talk about the vectors in a data frame as "columns", even though they have no true orientation. 

The tidyverse has strong opinions about how data frames should best be laid out, described as the ["tidy data" format](https://tidyr.tidyverse.org/). For now, we just need to pay attention to the constraint that the vectors in a data frame - the "colums" - can only contain one data type, but that type can vary between columns.

R has a number of built in data frames we can use for examples. Here's `mtcars`, a data frame about car performance data and specifications.

```{r}
head(mtcars)
```

We access elements of data.frames just the same way we access elements of any other list.

```{r}
efficiency <- mtcars[[4]] / mtcars[[3]]
# Take the 4th column of mtcars and divide it by the 3rd column of mtcars
# then assign that to `efficiency`
mean(efficiency)
```
This is pretty unclear, though. Not very reproducible, in the broader sense. A much better way to access data frame columns is by name. Basically you should only ever do this.
```{r}
efficiency <- mtcars$hp / mtcars$disp
```

# Other object classes
There are a million but they're mostly versions of lists. We won't worry about them now.

# A note on style

Code *style* is an important part of reproducibility, even if style by definition doesn't affect code execution (in which case it would be syntax, not style). Nevertheless, code written in bad style is hard to read, and therefore hard to evaluate for errors. I like to use RStudio's style guide [here](https://style.tidyverse.org/). I also make a slight modification:

* Vectors and matrices are always named using `.` to separate words: e.g. `named.vec`, `std.dev`, etc
* Anything else: functions, lists, data frames, etc., uses `_` to separate words: `named_list`, `my_data`, etc. Because tidyverse emphasizes using data frames, this is a quick way of keeping track of whether we are talking about a data frame (`_`) or a column within that data frame (`.`)

