---
title: "Lecture 1 v 2: Basics of R"
output: github_document
---

# R Resources

* [R for Data Science](https://r4ds.had.co.nz/)
* [tidyverse](https://www.tidyverse.org)
* [RStudio Cheat Sheets](https://www.rstudio.com/resources/cheatsheets/)

# Packages

```{r, messages=FALSE}
# install.packages(tidyverse) # gets pacakges from CRAN repository
library(tidyverse) # loads the package
```

Variable assignment and types

```{r}
a <- "tree" # a = "tree"
b <- 7
16 -> c # don't do this
```

We can do stuff with variables.

```{r}
print(b + c)
```

## Vectors and vectorization

Rules:

* Vecgtors are one-dimensional and ordered.
* Vectors can be any length (including 0 length).
* Vector elements an be names (but they aren't required to). We can access vector elements by position or by name
* Vectors must contain the same data type: numeric, character, logical (TRUE/FALSE), but *not* a mix of those.

```{r}
vec.1 <- c(1, 2, 3, 4, 5) # c stands for "concatenate"
vec.2 <- c(6, 5, 4, 3, 2)
vec.1 + vec.2
```
Use vectorization whenever possible! But be careful of vector lengths and recycling

```{r}
vec.2 + 3
```

```{r}
vec.2 + c(1, 2, 3)
```
 Note that we can name vector elements:
```{r}
named.vec <- c(a = 1, b = 2, c = 3)
```
 
